Metal Mixing and Ejection in Dwarf Galaxies is Dependent on Nucleosynthetic Source

### 2018-10-29 main.tex
removing textbfs for final version

One key aspect of this work has been a significant effort to characterize the number of independent dimensions accessible by the multi-dimensional chemical abundances observed in these studies \citep[e.g][]{Freeman2002,Ting2012, Hogg2016,Jofre2017,Price-Jones2018}. Thus far, these studies have remained uninformed by hydrodynamics simulations and, conversely, these results cannot yet be used to constrain simulations. Doing so requires the kinds of high resolution, galaxy-scale multi-element chemodynamical simulations that have only become feasible in recent years. \textbf{Recent work has suggested that low-mass dwarf galaxies are perhaps the best regime to begin understanding these processes \citep[e.g.][]{Bland-Hawthorn2010a,Bland-Hawthorn2010b,Karlsson2012,Webster2016}. This is advantageous, as their small physical size and low star formation rates makes conducting high resolution, hydrodynamics simulations of these systems computationally feasible.}

@@

\paragraph{Initial Conditions} Our dwarf galaxy is initialized to approximate, but not reproduce, the $z = 0$ properties of an ultrafaint dwarf galaxy (UFD) as informed by the observed properties of Leo P \citep[see ][]{Giovanelli2013,McQuinn2015a,McQuinn2015}. We initialize a M$_{\rm gas} = 1.8 \times 10^{6}$~M$_{\odot}$ disk as an exponential profile with a metal mass fraction of $Z = 4.3 \times 10^{-4}$ centered on a static \cite{Burkert1995} dark matter potential with M$_{\rm vir} = 2.5 \times 10^{9}$~M$_{\odot}$. The gas scale radius and scale height are set to 250~pc and 100~pc respectively, with a maximum radial extent of 600~pc. Both the gas temperatures and velocities are set iteratively to enforce initial hydrostatic equilibrium. The galaxy contains no initial background stellar population, limiting the number of Typa Ia SNe in our model. We include initial SN driving to limit the transient burst of star formation from the initial cooling and collapse of the galaxy. \textbf{These SNe are randomly distributed with the same radial and vertical scale heights as the gas disk at a rate of 0.4~Myr$^{-1}$; this corresponds to the current SFR of Leo P, $\sim 4\times 10^{-4}$~M$_{\odot}$~yr$^{-1}$ \citep{McQuinn2015a}, assuming 1 SNe per 100 M$_{\odot}$ of star formation. The metal yields from these SNe are set to the mean ISM abundances, and thus do not contribute to the chemical evolution of the galaxy.} Although the total metallicity field is initially non-zero, we only track the self-consistently produced metal enrichment for each individual metal species in our simulation, setting their initial abundances to zero. Our analysis is based on the evolution of this galaxy during the first 500~Myr after the formation of the first star particle.

@@

The log-normal distribution is found often in nature, generally describing multiplicative processes with non-negative values that grow with time. In astrophysics, for example, the log-normal distribution can be used to describe the time evolution of the star formation rate density \citep[see ][]{Gladders2013,Abramson2016,Diemer2017}. In addition, as expected from analytic theory \citep{Vazquez-Semadeni1994}, isothermal turbulence gives rise to log-normal density probability distribution functions \citep[PDFs;][]{Padoan1997, Passot1998, Ostriker1999,PadoanNordlund2002,KrumholzMcKee2005,Federrath2008}. Although these PDFs are only log-normal in simulations containing a more realistic, multi-phase ISM \citep{Scalo1998} if the disk is very stable \citep{WadaNorman2007}, individual phases within the ISM do exhibit some log-normality \citep{Tasker2009, Tasker2011,Joung2009,PriceFederrathBrunt2011, HopkinsQuataertMurray2012}. It has been shown that the 3D density PDF and the column density PDF, in both simulations and observations, have a characteristic shape \textbf{\citep{Vazquez-Semadeni1994,Burkhart2009, FederrathKlessen2013, Collins2012, Myers2015, Burkhart2017, Chen2018}}. This includes a log-normal component, generated by multiplicative processes---in this case shocks and the turbulent cascade in the ISM---and a power-law component at high densities arising from the additive combination of individual, self-gravitating cloud structures .

@@

Unsurprisingly the diffuse HIM \textbf{is the most metal rich phase}, as it is comprised predominately of metal enriched SN ejecta. Clearly this leads to long power-law tails towards high metal fractions for each species, with a very narrow, poorly defined peak at low metal fractions. Generally, in colder gas the metal PDFs become less enriched with broader, low-metal-fraction components and steeper, power-law tails. Although the extended power-law tail of the HIM leads to a large range of metal fractions, the HIM represents very little mass, and this tail represents recent, un-mixed enrichment. The comparatively narrow width in the log-normal regimes of the CNM is perhaps surprising. Although this gas is at lower metal fraction than the WNM and WIM, it would appear that it is more well mixed. This runs counter to the idea that mixing times should be long in colder gas, unless (as we argue) mixing first proceeds rapidly in the hot phases before mixing in with cold, disconnected structures across the galaxy.

Although these trends across phases hold for all elements, there are qualitative differences between elements, particularly between those ejected predominantly by AGB winds (e.g. Ba and N) and those ejected by SNe (e.g. O and Mg). In all phases, except the HIM, the AGB-wind elements have broader distributions that are less well described by our adopted $p(Z)$ than the metals dominated by SN enrichment. The power-law component of N and Ba is generally shallower than all other metals across each phase (except the HIM), particularly in the CNM. Ba and N do not show significant differences among the rest of the metals in the HIM, though this is likely because the Ba and N present in this phase is dominated by the Ba and N included in SNe yields. Again these differences between yield sources could be driven both by differences in their enrichment timescales and therefore differences in the typical ISM environment each event encounters, and the differences in energetics between AGB winds and SNe. \textbf{AGB wind elements enrich the WNM and WIM directly, rather than the HIM. This leads to longer mixing timescales and broader PDFs for these elements.}

@@

\subsection{Individual Stars vs. Averaged Yields}
\label{sec:SSP yields}
\textbf{Typical models for chemical enrichment in galaxy-scale simulations apply some form of IMF-averaged nucleosynthetic yields. Unfortunately, it is beyond the scope of this work to investigate how our results might change when adopting an IMF-averaged yield model. However, to what degree stochastic IMF sampling, mass-dependent nucleosynthetic yields, and the dynamical decoupling of individual enrichment sources play in galactic chemical evolution are valuable questions to address in future research. We speculate that an IMF-averaged enrichment model will only capture the differences between enrichment sources seen here if the model: 1) captures a multi-phase, turbulent ISM, 2) accounts for the energetic differences between yields from different enrichment sources, and 3) accounts for the time-delay between different enrichment sources. Models which average over an entire stellar population and ignore these effects may be unable to reproduce our results.}

>>>>

One key aspect of this work has been a significant effort to characterize the number of independent dimensions accessible by the multi-dimensional chemical abundances observed in these studies \citep[e.g][]{Freeman2002,Ting2012, Hogg2016,Jofre2017,Price-Jones2018}. Thus far, these studies have remained uninformed by hydrodynamics simulations and, conversely, these results cannot yet be used to constrain simulations. Doing so requires the kinds of high resolution, galaxy-scale multi-element chemodynamical simulations that have only become feasible in recent years. Recent work has suggested that low-mass dwarf galaxies are perhaps the best regime to begin understanding these processes \citep[e.g.][]{Bland-Hawthorn2010a,Bland-Hawthorn2010b,Karlsson2012,Webster2016}. This is advantageous, as their small physical size and low star formation rates makes conducting high resolution, hydrodynamics simulations of these systems computationally feasible.

@@

\paragraph{Initial Conditions} Our dwarf galaxy is initialized to approximate, but not reproduce, the $z = 0$ properties of an ultrafaint dwarf galaxy (UFD) as informed by the observed properties of Leo P \citep[see ][]{Giovanelli2013,McQuinn2015a,McQuinn2015}. We initialize a M$_{\rm gas} = 1.8 \times 10^{6}$~M$_{\odot}$ disk as an exponential profile with a metal mass fraction of $Z = 4.3 \times 10^{-4}$ centered on a static \cite{Burkert1995} dark matter potential with M$_{\rm vir} = 2.5 \times 10^{9}$~M$_{\odot}$. The gas scale radius and scale height are set to 250~pc and 100~pc respectively, with a maximum radial extent of 600~pc. Both the gas temperatures and velocities are set iteratively to enforce initial hydrostatic equilibrium. The galaxy contains no initial background stellar population, limiting the number of Typa Ia SNe in our model. We include initial SN driving to limit the transient burst of star formation from the initial cooling and collapse of the galaxy. These SNe are randomly distributed with the same radial and vertical scale heights as the gas disk at a rate of 0.4~Myr$^{-1}$; this corresponds to the current SFR of Leo P, $\sim 4\times 10^{-4}$~M$_{\odot}$~yr$^{-1}$ \citep{McQuinn2015a}, assuming 1 SNe per 100 M$_{\odot}$ of star formation. The metal yields from these SNe are set to the mean ISM abundances, and thus do not contribute to the chemical evolution of the galaxy. Although the total metallicity field is initially non-zero, we only track the self-consistently produced metal enrichment for each individual metal species in our simulation, setting their initial abundances to zero. Our analysis is based on the evolution of this galaxy during the first 500~Myr after the formation of the first star particle.

@@

The log-normal distribution is found often in nature, generally describing multiplicative processes with non-negative values that grow with time. In astrophysics, for example, the log-normal distribution can be used to describe the time evolution of the star formation rate density \citep[see ][]{Gladders2013,Abramson2016,Diemer2017}. In addition, as expected from analytic theory \citep{Vazquez-Semadeni1994}, isothermal turbulence gives rise to log-normal density probability distribution functions \citep[PDFs;][]{Padoan1997, Passot1998, Ostriker1999,PadoanNordlund2002,KrumholzMcKee2005,Federrath2008}. Although these PDFs are only log-normal in simulations containing a more realistic, multi-phase ISM \citep{Scalo1998} if the disk is very stable \citep{WadaNorman2007}, individual phases within the ISM do exhibit some log-normality \citep{Tasker2009, Tasker2011,Joung2009,PriceFederrathBrunt2011, HopkinsQuataertMurray2012}. It has been shown that the 3D density PDF and the column density PDF, in both simulations and observations, have a characteristic shape \citep{Vazquez-Semadeni1994,Burkhart2009, FederrathKlessen2013, Collins2012, Myers2015, Burkhart2017, Chen2018}. This includes a log-normal component, generated by multiplicative processes---in this case shocks and the turbulent cascade in the ISM---and a power-law component at high densities arising from the additive combination of individual, self-gravitating cloud structures .

@@

Unsurprisingly the diffuse HIM is the most metal rich phase, as it is comprised predominately of metal enriched SN ejecta. Clearly this leads to long power-law tails towards high metal fractions for each species, with a very narrow, poorly defined peak at low metal fractions. Generally, in colder gas the metal PDFs become less enriched with broader, low-metal-fraction components and steeper, power-law tails. Although the extended power-law tail of the HIM leads to a large range of metal fractions, the HIM represents very little mass, and this tail represents recent, un-mixed enrichment. The comparatively narrow width in the log-normal regimes of the CNM is perhaps surprising. Although this gas is at lower metal fraction than the WNM and WIM, it would appear that it is more well mixed. This runs counter to the idea that mixing times should be long in colder gas, unless (as we argue) mixing first proceeds rapidly in the hot phases before mixing in with cold, disconnected structures across the galaxy.

Although these trends across phases hold for all elements, there are qualitative differences between elements, particularly between those ejected predominantly by AGB winds (e.g. Ba and N) and those ejected by SNe (e.g. O and Mg). In all phases, except the HIM, the AGB-wind elements have broader distributions that are less well described by our adopted $p(Z)$ than the metals dominated by SN enrichment. The power-law component of N and Ba is generally shallower than all other metals across each phase (except the HIM), particularly in the CNM. Ba and N do not show significant differences among the rest of the metals in the HIM, though this is likely because the Ba and N present in this phase is dominated by the Ba and N included in SNe yields. Again these differences between yield sources could be driven both by differences in their enrichment timescales and therefore differences in the typical ISM environment each event encounters, and the differences in energetics between AGB winds and SNe. AGB wind elements enrich the WNM and WIM directly, rather than the HIM. This leads to longer mixing timescales and broader PDFs for these elements.

@@

\subsection{Individual Stars vs. Averaged Yields}
\label{sec:SSP yields}
Typical models for chemical enrichment in galaxy-scale simulations apply some form of IMF-averaged nucleosynthetic yields. Unfortunately, it is beyond the scope of this work to investigate how our results might change when adopting an IMF-averaged yield model. However, to what degree stochastic IMF sampling, mass-dependent nucleosynthetic yields, and the dynamical decoupling of individual enrichment sources play in galactic chemical evolution are valuable questions to address in future research. We speculate that an IMF-averaged enrichment model will only capture the differences between enrichment sources seen here if the model: 1) captures a multi-phase, turbulent ISM, 2) accounts for the energetic differences between yields from different enrichment sources, and 3) accounts for the time-delay between different enrichment sources. Models which average over an entire stellar population and ignore these effects may be unable to reproduce our results.
